---
title: "13_Thesis_Figures"
author: "noor ahmed"
date: "2026-07-19"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "...") # make own directory
```

```{r load_libraries}
library(tidyverse)
library(ggplot2)
library(ggrepel)
library(pheatmap)

dir.create("figures", showWarnings = FALSE)
```

```{r figure1_qc}
# Figure 1 - QC boxplot and PCA
# already saved in QC/Microarray/ - just copy to figures/
file.copy("QC/Microarray/GSE3307_normalised_boxplot.png",
          "figures/Figure1A_QC_boxplot.png", overwrite = TRUE)
file.copy("QC/Microarray/GSE3307_normalised_PCA.png",
          "figures/Figure1B_QC_PCA.png", overwrite = TRUE)
cat("Figure 1 copied\n")
```

```{r figure2_volcano}
# Figure 2 - Volcano plots
# already saved - copy to figures/
file.copy("Volcano_Plots/GSE3307_volcano.png",
          "figures/Figure2A_volcano_ALS.png", overwrite = TRUE)
file.copy("Volcano_Plots/GSE140089_volcano.png",
          "figures/Figure2B_volcano_PD.png", overwrite = TRUE)
cat("Figure 2 copied\n")
```

```{r figure3_upset}
# Figure 3 - UpSet plot cross-disease overlap
file.copy("Consistency/CrossDisease_UpSet.png",
          "figures/Figure3_UpSet.png", overwrite = TRUE)
cat("Figure 3 copied\n")
```

```{r figure4_go_dotplot}
# Figure 4 - GO pathway dotplot
file.copy("Pathway/ALS_PD_shared_GO_simplified_dotplot.png",
          "figures/Figure4_GO_dotplot.png", overwrite = TRUE)
cat("Figure 4 copied\n")
```

```{r figure5_gsea}
# Figure 5 - GSEA dotplot
file.copy("GSEA/GSE3307_GSEA_dotplot.png",
          "figures/Figure5_GSEA.png", overwrite = TRUE)
cat("Figure 5 copied\n")
```

```{r figure6_muscle_cns_heatmap}
# Figure 6 - Muscle vs CNS heatmap
# already generated in 12_CNS_Comparison.Rmd
# if not there, re-run that chunk here
file.copy("figures/Muscle_vs_CNS_heatmap.png",
          "figures/Figure6_CNS_heatmap.png", overwrite = TRUE)
cat("Figure 6 copied\n")
```

```{r figure7_muscle_cns_pattern}
# Figure 7 - Muscle vs CNS pattern summary
file.copy("figures/Muscle_vs_CNS_pattern_summary.png",
          "figures/Figure7_CNS_pattern.png", overwrite = TRUE)
cat("Figure 7 copied\n")
```

```{r figure8_biomarker_tiers}
# Figure 8 - Biomarker tier summary
final_table <- read.csv("Biomarkers/Final_Tiered_Biomarker_Candidates.csv",
                         stringsAsFactors = FALSE)

plot_data <- final_table %>%
  head(20) %>%
  mutate(
    gene = factor(gene, levels = rev(gene)),
    tier_short = case_when(
      grepl("Tier 1", biomarker_tier) ~ "Tier 1: Pan-ND and ALS-specific",
      grepl("Tier 2", biomarker_tier) ~ "Tier 2: ALS-specific",
      grepl("Tier 3", biomarker_tier) ~ "Tier 3: Pan-ND, partially specific",
      grepl("Tier 4", biomarker_tier) ~ "Tier 4: Partially specific",
      grepl("Tier 5", biomarker_tier) ~ "Tier 5: Pan-ND but generic",
      grepl("Tier 6", biomarker_tier) ~ "Tier 6: Generic muscle disease",
      TRUE ~ biomarker_tier
    )
  )

tier_colours <- c(
  "Tier 1: Pan-ND and ALS-specific"    = "#A50026",
  "Tier 2: ALS-specific"               = "#D6604D",
  "Tier 3: Pan-ND, partially specific" = "#F4A582",
  "Tier 4: Partially specific"         = "#92C5DE",
  "Tier 5: Pan-ND but generic"         = "#4393C3",
  "Tier 6: Generic muscle disease"     = "#2166AC"
)

p <- ggplot(plot_data, aes(x = score_updated, y = gene,
                             fill = tier_short)) +
  geom_bar(stat = "identity", colour = "white", linewidth = 0.3) +
  geom_text(aes(label = round(score_updated, 1)),
            hjust = -0.2, size = 3.5) +
  scale_fill_manual(values = tier_colours, name = "Biomarker Tier") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "bottom",
    legend.text = element_text(size = 9),
    panel.grid.major.y = element_blank(),
    axis.text.y = element_text(size = 10, face = "bold")
  ) +
  labs(
    title = "Biomarker Candidate Prioritisation — Top 20 Genes",
    subtitle = "Scored on consistency, effect size, pathway relevance,\ncross-disease evidence and specificity",
    x = "Weighted Score",
    y = "Gene"
  ) +
  xlim(0, max(plot_data$score_updated) * 1.15)

ggsave("figures/Figure8_Biomarker_tiers.png",
       plot = p, width = 10, height = 8, dpi = 300)

cat("Figure 8 saved\n")
```


No genes reached Tier 1 — defined as candidates that were both nominally significant across all three diseases and ALS-specific relative to non-neurodegenerative muscle diseases. This indicates that no single gene simultaneously demonstrated pan-neurodegenerative relevance and neurodegeneration-specific expression, suggesting that cross-disease conservation and disease specificity represent competing rather than complementary properties in the skeletal muscle transcriptional response to neurodegeneration. The strongest candidates therefore represent a trade-off: MGST3 and RNASEL (Tier 3) show broader disease relevance across ALS, PD and VCP-myopathy but partial specificity, while the eight Tier 2 genes are ALS-specific but not conserved across diseases.


Tier 1 — Pan-ND and ALS-specific
Significant across multiple neurodegenerative diseases and not found in any other muscle disease — the strongest possible biomarker candidate.

Tier 2 — ALS-specific only
Significant in ALS but not in other muscle diseases — specific to ALS but not shared across neurodegeneration.

Tier 3 — Pan-ND, partially specific
Significant across multiple neurodegenerative diseases and found in only a few other muscle diseases — strong cross-disease candidates with reasonable specificity.

Tier 4 — Partially specific
Significant in one or two neurodegenerative diseases and found in a small number of other muscle diseases — moderate specificity.

Tier 5 — Pan-ND but generic
Significant across multiple neurodegenerative diseases but also found in many other muscle diseases — broad but not specific to neurodegeneration.

Tier 6 — Generic muscle disease
Significant in neurodegeneration but also found across most other muscle diseases — reflects general muscle pathology rather than neurodegeneration specifically.


```{r check_all_figures}
# confirm all 8 figures are present
cat("All thesis figures:\n")
list.files("figures", pattern = "Figure")
```

```{r}
library(tidyverse)
library(ggplot2)

# load the comparator DE results
als_vs_jdm  <- read.csv("Specificity/ALS_vs_JDM_DE.csv",  stringsAsFactors = FALSE)
als_vs_spg  <- read.csv("Specificity/ALS_vs_SPG_DE.csv",  stringsAsFactors = FALSE)
als_vs_dmd  <- read.csv("Specificity/ALS_vs_DMD_DE.csv",  stringsAsFactors = FALSE)
als_vs_fshd <- read.csv("Specificity/ALS_vs_FSHD_DE.csv", stringsAsFactors = FALSE)
als_vs_aqm  <- read.csv("Specificity/ALS_vs_AQM_DE.csv",  stringsAsFactors = FALSE)

top_genes <- c("FOSL1", "TPRG1", "PILRB", "ACSS1", "RFLNA",
                "SLC40A1", "B3GLCT", "PDE7A", "MGST3", "RNASEL",
                "TNNC1", "MYH7", "TNNT1", "TNNI1", "MYOZ2",
                "SOCS3", "KCNN3", "TPM3", "PPP3CC", "PDLIM5")

# build specificity matrix for top candidates
spec_data <- data.frame(
  gene        = top_genes,
  sig_vs_JDM  = top_genes %in% filter(als_vs_jdm,  adj.P.Val < 0.05)$gene,
  sig_vs_SPG  = top_genes %in% filter(als_vs_spg,  adj.P.Val < 0.05)$gene,
  sig_vs_DMD  = top_genes %in% filter(als_vs_dmd,  adj.P.Val < 0.05)$gene,
  sig_vs_FSHD = top_genes %in% filter(als_vs_fshd, adj.P.Val < 0.05)$gene,
  sig_vs_AQM  = top_genes %in% filter(als_vs_aqm,  adj.P.Val < 0.05)$gene
) %>%
  mutate(
    n_comparators = sig_vs_JDM + sig_vs_SPG + sig_vs_DMD +
                    sig_vs_FSHD + sig_vs_AQM,
    tier = case_when(
      n_comparators == 0 ~ "ALS-specific (Tier 2)",
      n_comparators <= 2 ~ "Partially specific (Tier 3/4)",
      n_comparators >= 3 ~ "Generic muscle disease (Tier 5/6)"
    )
  )

cat("Specificity breakdown:\n")
print(spec_data %>% dplyr::select(gene, n_comparators, tier))

# reshape for heatmap
spec_long <- spec_data %>%
  dplyr::select(gene, sig_vs_JDM, sig_vs_SPG,
                sig_vs_DMD, sig_vs_FSHD, sig_vs_AQM) %>%
  pivot_longer(-gene,
               names_to  = "comparator",
               values_to = "significant") %>%
  mutate(
    comparator = gsub("sig_vs_", "", comparator),
    gene = factor(gene, levels = rev(top_genes))
  )

# add tier annotation for gene labels
tier_labels <- spec_data %>%
  dplyr::select(gene, tier) %>%
  mutate(gene = factor(gene, levels = rev(top_genes)))

p <- ggplot(spec_long, aes(x = comparator, y = gene,
                             fill = significant)) +
  geom_tile(colour = "white", linewidth = 0.8) +
  scale_fill_manual(
    values = c("TRUE"  = "#D6604D",
               "FALSE" = "#F7F7F7"),
    labels = c("TRUE"  = "Significant (padj < 0.05)",
               "FALSE" = "Not significant"),
    name = ""
  ) +
  theme_bw(base_size = 12) +
  theme(
    panel.grid    = element_blank(),
    axis.text.x   = element_text(size = 11, face = "bold"),
    axis.text.y   = element_text(size = 10),
    legend.position = "bottom",
    plot.title    = element_text(face = "bold", size = 13),
    plot.subtitle = element_text(size = 10, colour = "grey40")
  ) +
  labs(
    title    = "ALS Biomarker Candidate Specificity",
    subtitle = "Red = also significant when ALS compared against non-neurodegenerative muscle disease",
    x        = "Muscle disease comparator",
    y        = "Gene"
  )

ggsave("figures/Figure10_Specificity_heatmap.png",
       plot = p, width = 8, height = 10, dpi = 300)

cat("Figure 10 saved\n")
```

